The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-30 article_final/main.tex
fixed Nesterov formulation

We then compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT and GPT-2 Medium on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where \texttt{sign\_lr\_coeff} $\eta$ is a scaling coefficient specific to SignSGD.

@@

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [5\mathrm{e}{-5},\, 2\mathrm{e}{-4}]$ with step $1\mathrm{e}{-5}$.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 0.0003/0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

@@

\subsection{GPT-2 Medium Speedrun}
We scale from NanoGPT to GPT-2 Medium. After testing for 5960 steps we get 2.9215 loss for F-Muon and 2.9235 for S-Muon, while for Muon it is 2.9198 (the speedrun threshold is 2.92).

@@

Common important drawback of the analyses is the consideration of the convergence by the norm of the gradient. As we showed in our experiments on CIFAR \cref{sec:cifar_underhood}, the gradient norm may decrease only by a factor of ten even when the accuracy reaches 100\%.

@@

\section{Under the hood of CNN on CIFAR-10}
\label{sec:cifar_underhood}

>>>>

We then compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where $\eta$ is a scaling coefficient specific to SignSGD.

@@

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [0.00005, 0.00020]$ with step $0.00001$ for SignSGD.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 0.0003 for F-Muon and 0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.28. As illustrated in \figureref{fig:gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

@@

\subsection{GPT-2 Medium Speedrun}
We scale from NanoGPT to GPT-2 Medium. After testing for 5960 steps with F-Muon we get 2.9215 loss, while for Muon it is 2.9198. In ? steps, F-Muon reaches <2.92 loss, thus needing only ?\% more iterations than Muon.

@@

Common important drawback of the analyses is the consideration of the convergence by the norm of the gradient. As we showed in our experiments on CIFAR, the gradient norm may decrease only by a factor of ten, when the accuracy reaches 100\%.

@@

\section{Under the hood of CNN on CIFAR-10}
